Current state and prospects of R-packages for the design of experiments

### 2022-12-13 paper/arxiv/arxiv.Rmd
update arxiv

---
authors:
  - name: Emi Tanaka
    thanks: Corresponding author
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Dewi Amaliah
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
header-includes: |
  \usepackage{colortbl}
  \usepackage{bera}
  \usepackage{xcolor}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}
bibliography: ../../paper.bib
biblio-style: unsrt
#preamble: >
output: 
  rticles::arxiv_article:
    keep_tex: true
---

```{r, include = FALSE}
paper <- here::here("paper.Rmd")
meta <- rmarkdown::yaml_front_matter(paper)
```

>>>>

---
authors:
  - name: Emi Tanaka
    thanks: Corresponding author
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Dewi Amaliah
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
header-includes: |
  \usepackage{colortbl}
  \usepackage{bera}
  \usepackage{xcolor}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}
bibliography: paper.bib
biblio-style: unsrt
#preamble: >
output: 
  rticles::arxiv_article:
    keep_tex: true
---

```{r, include = FALSE}
file.copy(here::here("paper.bib"), here::here("paper/arxiv/paper.bib"))
file.copy(here::here("supp.pdf"), here::here("paper/arxiv/anc/supp.pdf"))
zip("anc", "anc/")

paper <- here::here("paper.Rmd")
meta <- rmarkdown::yaml_front_matter(paper)
```

### 2022-12-13 paper/arxiv/arxiv.tex
update arxiv

In
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 114 R packages for experimental design and analysis
of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, e.g.~in Python, only a handful of packages
that generate design of experiment exist (namely \texttt{pyDOE},
\texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter} and
\texttt{GPdoemd}) with limited type of designs. Thus, the study of DoE
packages, based on quantitative and qualitative data, can give us an
objective view to the state of the current experimental designs in
practice.

@@

All CRAN packages have a title, description, package connections
(suggests, depends and imports of other packages) and other
meta-information in the DESCRIPTION file. We use the text data from the
title and description (as accessed in 2022-05-23) in Section
\ref{topics}.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN
relevant to the corresponding topic. There are a total 37 CRAN task
views. Table S1 in the Supplementary Materials list the full available
topics from the \texttt{ctv} package (Zeileis 2005). The list of
packages in each CRAN task views (as of 2022-05-25) are used to contrast
the characteristics of DoE packages in Sections \ref{popular} and
\ref{silo}.

@@

There are at least 50 DoE packages since 2013 but most of the downloads
are concentrated in just a handful of packages. For example, Figure
\ref{fig:plot-lorenz} shows a Lorenz curve (Lorenz 1905) for the total
package downloads in 2021 of 113 DoE packages (first released prior to
2021); we can see from Figure \ref{fig:plot-lorenz} that bottom 90\% of
DoE packages (in terms of total download count in 2021) only share about
27\% of total downloads across all DoE packages -- in another words,
73\% of the total downloads are due to 11 packages (10\% of the DoE
packages).

@@

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region over the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages while a value of 100\% indicates maximal inequality (all
downloads are due to one package). In Figure \ref{fig:download-share},
we see that the distributions of the package downloads each year have a
heavy right tail with the Gini index ranging from 42.6\% to 74.1\%
across the years 2013 to 2021, indicating that there is a high level of
inequality of package downloads, particularly with more pronounced
inequality in the last 6 years.

@@

While in absolute terms the Gini index is high for
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view (42.6\% to 74.1\%), the inequality is not as severe as other
CRAN task views as shown in Figure \ref{fig:fig-gini-all-ctvs}. We can
see in Figure \ref{fig:fig-gini-all-ctvs} that the Gini index is
generally increasing for DoE packages (as is generally the case for
other CRAN task views as shown in Figure S1 in the Supplementary
Materials) but most other CRAN task views have a Gini index of over
75\%. This suggests that other CRAN task views may have dominant
standards and in comparison to other topics, there are more
\textbf{\emph{diverse approaches to designing experiments}} and thus no
single DoE package is dominant. This observation however doesn't take
into account other approaches to generate experimental designs, like say
the proprietary software, CycDesignN (Whittaker, Williams, and John
2022), which may be widely used.

@@

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has the lowest average number of contributors among all 37
CRAN task views. In addition, we can also see in Figure
\ref{fig:ctv-summ-plot} that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection of DoE packages with other DoE packages is shown in
Figure \ref{fig:plot-doe-network}. These observations suggest that the
\textbf{\emph{field of experimental design is one of the least
collaborative field}} and package development generally occur in silos.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BHH2}{}}%
Barrios, Ernesto. 2016. \emph{Bhh2: Useful Functions for Box, Hunter and
Hunter II}. \url{https://CRAN.R-project.org/package=BHH2}.

@@

\bibliographystyle{unsrt}
\bibliography{../../paper.bib}

>>>>

In
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 105 R packages for experimental design and analysis
of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, e.g.~in Python, only a handful of packages
that generate design of experiment exist (namely \texttt{pyDOE},
\texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter} and
\texttt{GPdoemd}) with limited type of designs. Thus, the study of DoE
packages, based on quantitative and qualitative data, can give us an
objective view to the state of the current experimental designs in
practice.

@@

All CRAN packages have a title, description, package connections
(suggests, depends and imports of other packages) and other
meta-information in the DESCRIPTION file. We use the text data from the
title and description (as accessed in 2022-12-12) in Section
\ref{topics}.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN
relevant to the corresponding topic. There are a total 39 CRAN task
views. Table S1 in the Supplementary Materials list the full available
topics from the \texttt{ctv} package (Zeileis 2005). The list of
packages in each CRAN task views (as of 2022-12-12) are used to contrast
the characteristics of DoE packages in Sections \ref{popular} and
\ref{silo}.

@@

There are at least 50 DoE packages since 2013 but most of the downloads
are concentrated in just a handful of packages. For example, Figure
\ref{fig:plot-lorenz} shows a Lorenz curve (Lorenz 1905) for the total
package downloads in 2021 of 102 DoE packages (first released prior to
2021); we can see from Figure \ref{fig:plot-lorenz} that bottom 90\% of
DoE packages (in terms of total download count in 2021) only share about
32\% of total downloads across all DoE packages -- in another words,
68\% of the total downloads are due to 10 packages (10\% of the DoE
packages).

@@

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region over the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages while a value of 100\% indicates maximal inequality (all
downloads are due to one package). In Figure \ref{fig:download-share},
we see that the distributions of the package downloads each year have a
heavy right tail with the Gini index ranging from 32.7\% to 69.1\%
across the years 2013 to 2021, indicating that there is a high level of
inequality of package downloads, particularly with more pronounced
inequality in the last 6 years.

@@

While in absolute terms the Gini index is high for
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view (32.7\% to 69.1\%), the inequality is not as severe as other
CRAN task views as shown in Figure \ref{fig:fig-gini-all-ctvs}. We can
see in Figure \ref{fig:fig-gini-all-ctvs} that the Gini index is
generally increasing for DoE packages (as is generally the case for
other CRAN task views as shown in Figure S1 in the Supplementary
Materials) but most other CRAN task views have a Gini index of over
75\%. This suggests that other CRAN task views may have dominant
standards and in comparison to other topics, there are more
\textbf{\emph{diverse approaches to designing experiments}} and thus no
single DoE package is dominant. This observation however doesn't take
into account other approaches to generate experimental designs, like say
the proprietary software, CycDesignN (Whittaker, Williams, and John
2022), which may be widely used.

@@

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has the lowest average number of contributors among all 39
CRAN task views. In addition, we can also see in Figure
\ref{fig:ctv-summ-plot} that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection of DoE packages with other DoE packages is shown in
Figure \ref{fig:plot-doe-network}. These observations suggest that the
\textbf{\emph{field of experimental design is one of the least
collaborative field}} and package development generally occur in silos.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BHH2}{}}%
Barrios, Ernesto. 2016. \emph{BHH2: Useful Functions for Box, Hunter and
Hunter II}. \url{https://CRAN.R-project.org/package=BHH2}.

@@

\bibliographystyle{unsrt}
\bibliography{paper.bib}
